Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-08-25 2024/OPTIMA/OPTIMA_2024_MCO.tex
fix last comments

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparative efficiency of machine learning models for enhancing algorithms in solving multiextremal multicriteria problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by by the Research and Education Mathematics Centre ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriteria problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

\textcolor[rgb]{0,0,1}{\textit{Remark 1.} The Lipschitz constant of the objective function is assumed a priori unknown; it is adaptively estimated during the search.
The value $M$ from (\ref{eq:10}) can be used in Lipschitz constant estimation.\\
\textit{Remark 2.} It follows from relationship between the Lipschitz and H{\" o}lder constants of the functions $\varphi(y)$ and $\varphi(y(x))$, as well as from the theorem of convergence of the global search algorithm (see \cite{Strongin2000}), that parameter $r$ from (\ref{eq:10}) should satisfy the necessary condition $r>2^{2-1/N}$. 
In the test experiments described in Section~\ref{sec:5} we used the value $r=3$, which ensured the correct solution of the entire series of problems. In applied optimization problems the value $r=3$ also ensures the solution in the most cases.
}

@@

For computation of $R_{ps}(i)$ we propose to apply the machine learning methods. It should be noted that in classical combinations of machine learning and optimization methods, the latter are used as an auxiliary internal techniques that improve the machine learning functioning. 
\textcolor[rgb]{0,0,1}{Other popular approaches use machine learning for auto-configuration (tuning the hyper-parameters) of optimization algorithms (see, e.g., \cite{Bezerra2020,Lopez2016} and the references given therein)}.
In our approach, on the contrary, machine learning methods are a part of the optimization algorithms enhancing their work.

@@

In each series of experiments for these optimization algorithms, the coefficients $\alpha$ from (\ref{eq:16}) changed in the range $[0.1, 0.9]$. Comparison with the algorithm MGSA that does not use machine learning methods (i.e. with $\alpha = 0$ in the characteristic (\ref{eq:16})) and with the algorithm ML\_MGSA\_Dist taking into account the distance to the separating hyperplane was carried out. The comparison was made on the base of the HV index characterizing quality of the obtained Pareto set approximation. 
\textcolor[rgb]{0,0,1}{The HyperVolume (HV) index corresponds to the volume of the region dominated by the Pareto set approximation and bounded by a given reference point \cite{ML_MCO_2023,Evtushenko2014}. The larger the hypervolume index, the better is the corresponding approximation.
}

@@

The following parameters are indicated in Table \ref{tab:1}: the kernel of the SVM model (column 2), parameter $\alpha$ (column 3), average values of the HV-index (column 4), the average number of trials spent by the algorithm (column 5), and the coefficient of decreasing the number of trials compared to ML\_MGSA (column 6). The factors $r$ and $\varepsilon$ from (\ref{eq:10}) and (\ref{eq:14}) were the same for all the algorithms.

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[t]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

Fig. \ref{fig:3} shows the change of the average HV-index as a function of the number of trials.
\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Comparison of HV-indices for the case of fixed weights.} 
\label{fig:3}
\end{figure}

Compared to MGSA, the other algorithms reduce significantly the number of iterations in conditions when HV-indices are similar. However, the iteration number when using the models with probabilities (see rows ``ML\_MGSA\_Prob (Fixed weights)'' of Table \ref{tab:1}) is greater than for the algorithm based on the separating hyperplane.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig4.png}
\caption{HV-indices comparison  in the case of adjusted weights} 
\label{fig:4}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of  HV-indices for methods with probabilities (\ref{eq:18})} 
\label{fig:5}
\end{figure}

@@

The results of the experiments allow us to conclude that the method ML\_MGSA\_Prob (Log Norm) in this series of experiments can significantly reduce the number of trials with a similar value of HV-index compared to ML\_MGSA\_Dist. The best results were achieved using models with Linear and RBF kernels, while the latter model demonstrates better quality than the version of the algorithm with the technique \cite{ML_MCO_2023} based on distances to the hyperplane.

@@

Acceleration of the algorithm (in terms of reducing the number of search trials required to evaluate \textcolor[rgb]{0,0,1}{the Pareto set approximation with a certain precision}) is achieved through the application of machine learning methods. Namely, when using the SVM method, trial points in the criteria space are classified (each point is/is not an efficient solution), on the basis of which a numerical estimate of the probability of belonging to the Pareto set is built. Polynomial and radial basis function kernels in SVM allows one to adequately approximate Pareto front of a complex structure.

Solving several series of MCO problems with complex multiextremal criteria has demonstrated the potential of the proposed approach to evaluate the set of efficient solutions. Further research aims at extending the proposed approach by considering other machine learning models (in particular, artificial neural networks) to better classify the trials and approximate the Pareto set.

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparative efficiency of machine learning models for enhancing algorithms in solving multiextremal multicriteria problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by the Research and Education Mathematics Center ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriteria problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

\textit{Remark 1.} The Lipschitz constant of the objective function is assumed a priori unknown; it is adaptively estimated during the search.
The value $M$ from (\ref{eq:10}) can be used in Lipschitz constant estimation.

\textit{Remark 2.} It follows from relationship between the Lipschitz and H{\" o}lder constants of the functions $\varphi(y)$ and $\varphi(y(x))$, as well as from the theorem of convergence of the global search algorithm (see \cite{Strongin2000}), that parameter $r$ from (\ref{eq:10}) should satisfy the necessary condition $r>2^{2-1/N}$. 
In the test experiments described in Section~\ref{sec:5} we used the value $r=3$, which ensured the correct solution of the entire series of problems. In applied optimization problems the value $r=3$ also ensures the solution in the most cases.

@@

For computation of $R_{ps}(i)$ we propose to apply the machine learning methods. It should be noted that in classical combinations of machine learning and optimization methods, the latter are used as an auxiliary internal techniques that improve the machine learning functioning. 
Other popular approaches use machine learning for auto-configuration (tuning the hyperparameters) of optimization algorithms (see, e.g., \cite{Bezerra2020,Lopez2016} and the references given therein).
In our approach, on the contrary, machine learning methods are a part of the optimization algorithms enhancing their work.

@@

In each series of experiments for these optimization algorithms, the coefficients $\alpha$ from (\ref{eq:16}) changed in the range $[0.1, 0.9]$. Comparison with the algorithm MGSA that does not use machine learning methods (i.e. with $\alpha = 0$ in the characteristic (\ref{eq:16})) and with the algorithm ML\_MGSA\_Dist taking into account the distance to the separating hyperplane was carried out. The comparison was made on the base of the HV index characterizing quality of the obtained Pareto set approximation. 
The HyperVolume (HV) index is a traditional indicator of quality when approximating a Pareto set \cite{ML_MCO_2023,Evtushenko2014}. It corresponds to the volume of the region dominated by the Pareto set approximation obtained in the course of numerical solving. The greater the HV index, the better the approximation.

@@

The following parameters are indicated in Table \ref{tab:1}: the kernel of the SVM model (column 2), parameter $\alpha$ (column 3), average values of the HV index (column 4), the average number of trials spent by the algorithm (column 5), and the coefficient of decreasing the number of trials compared to ML\_MGSA (column 6). The factors $r$ and $\varepsilon$ from (\ref{eq:10}) and (\ref{eq:14}) were the same for all the algorithms.

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[t]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV index} & \textbf{Avg. iter.} & \textbf{Speedup} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Comparison of HV indices for the case of fixed weights.} 
\label{fig:3}
\end{figure}

Fig. \ref{fig:3} shows the change of the average HV index as a function of the number of trials.
Compared to MGSA, the other algorithms reduce significantly the number of iterations in conditions when HV indices are similar. However, the iteration number when using the models with probabilities (see rows ``ML\_MGSA\_Prob (Fixed weights)'' of Table \ref{tab:1}) is greater than for the algorithm based on the separating hyperplane.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig4.png}
\caption{HV indices comparison  in the case of adjusted weights} 
\label{fig:4}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of  HV indices for methods with probabilities (\ref{eq:18})} 
\label{fig:5}
\end{figure}

@@

The results of the experiments allow us to conclude that the method ML\_MGSA\_Prob (Log Norm) in this series of experiments can significantly reduce the number of trials with a similar value of HV index compared to ML\_MGSA\_Dist. The best results were achieved using models with Linear and RBF kernels, while the latter model demonstrates better quality than the version of the algorithm with the technique \cite{ML_MCO_2023} based on distances to the hyperplane.

@@

Acceleration of the algorithm (in terms of reducing the number of search trials required to evaluate the Pareto set approximation with a certain precision) is achieved through the application of machine learning methods. Namely, when using the SVM method, trial points in the criteria space are classified (each point is/is not an efficient solution), on the basis of which a numerical estimate of the probability of belonging to the Pareto set is built. Polynomial and radial basis function kernels in SVM allows one to adequately approximate Pareto front of a complex structure.

Solving several series of MCO problems with complex multiextremal criteria has demonstrated the potential of the proposed approach to evaluate the set of efficient solutions. Further research aims at extending the proposed approach by considering other machine learning models (in particular, artificial neural networks) to better classify the trials and approximate the Pareto set. Additionally, we intend to expand the basis of computational experiment by using new test sets, for example, the benchmarks \cite{Deb2003} and considering the statistical analysis to estimate test results more qualitative.
